Remote sensing riverine gas exchange and estimating carbon efflux using SWOT observations

### 2021-02-19 manuscript_v1.Rmd
min obs slope bug

##Key Points (85 characters each)
- We introduce the BIKER algorithm to remotely sense riverine gas exchange using the SWOT satellite
- BIKER replicates estimates of $CO_2$ fluxes when coupled with in situ concentration data
- Expected SWOT measurement errors exert a trivial influence on BIKER's estimates
- BIKER's accuracy is limited by process-level uncertainties and not by its algorithmic setup

@@

$$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})  \text{(5)}$$

@@

##4 Discussion
###4.1 Towards remote sensing global spatiotemporal dynamics of $k_{600}$
To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g. @alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @schelkerCO2EvasionSteep2016a], 3) discharge [e.g. @robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998], or 4) oxygen diel curves (where gas exchange is still assumed constant in time- e.g. @applingOvercomingEquifinalityLeveraging2018; @demarsStreamMetabolismOpen2015). However, the spatiotemporal dynamics of gas exchange are still weakly constrained. A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012]. [@wallinSpatiotemporalVariabilityGas2011] performed a preliminary analysis in northern Sweden relating river specific (‘at-a-station’) temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed. This limited knowledge of large-scale $k_{600}$ spatiotemporal dynamics is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also a lack of measurements. Authors have argued that the key to explaining the large residual variation in upscaling models is to explore at-a-station temporal variability in $k_{600}$ [@hallGasExchangeStreams2020].

@@

###4.2 Estimating bulk carbon efflux from SWOT
Section 3.3 confirms that BIKER is successful at 1) reproducing $FCO_2$ without any in situ river information for $k_{600}$ (Figure 7) and 2) reproducing the bulk carbon efflux without any in situ information for $k_{600}$ (Figure 5.8). This encouraging result has two main implications for future work. First, we can likely couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions, given that the temporal resolution of $k_600$ has arguably lagged behind recent advances in gas concentration data **get citations from Kelly**. BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope following recent work doing the same using the [@hagemannBAMBayesianAMHGManning2017a] RSQ algorithm [@harlanDischargeEstimationDense].

Secondly, it is important to stress that, unlike BIKER, the HG models in Figure 8 rely on an in situ streamgauge. This means that Figure 8 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease, though this decrease would not be too much due to the generally good performance of those data. Thus, these results suggest that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our bulk efflux understandings at gauged sites too. Future work should systematically quantify prediction error from using both global-scope HG and modeled discharge, like is the default workflow used in recent upscaling studies [e.g. @lauerwaldSpatialPatternsCO22015, @horgbyUnexpectedLargeEvasion2019]. **Iassume Shaoda's paper will still be in review and can't cite here **

@@

###4.2 $K_{600}$ estimate uncertainity from physical processes verus from RS
Throughout the BIKER validation, we have assumed no parameter uncertainity in the upscaling parameters $a_i$ and $b_i$. We have shown that BIKER estimate uncertainty is significantly less than $a_i$ and $b_i$ uncertainity (Figures 4 and 5). Therefore, BIKER provides similar estimates of $k_{600}$ to those produced by current upscaling models and most of the total uncertainity stems from the upscaling model itself (Figure 4). Therefore, we argue that BIKER's estimate uncertainity is limited by current process-level knowledge of $k_{600}$ upscaling and not by BIKER or SWOT, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the processes governing gas exchange from SWOT-observable rivers. Validating the BIKER algorithm has provided a unique lense to view $k_600$ upscaling through, and in that context we next explore some considerations for future work on $k_{600}$ upscaling in SWOT-observable rivers.

####4.2.1 Gas exhange under uniform and non-uniform flow conditions
Section 2.2 highlights that most upscaling studies to date have assumed uniform flow conditions (i.e. $S_h \neq S_0$) in order to train upscaling models using readily available slope data. However, the first-principles model previously used by [@ulsethDistinctAirWater2019a] and [@raymondScalingGasTransfer2012a] to define *eD* [@tsivoglouTracerMeasurementReaeration1976] does not make this simplifying assumption. Therefore, it is an open research question whether parameterizing $k_{600}$ upscaling models via $S_h$ can account for some of the unexplained residual variation in current upscaling models [@hallGasExchangeStreams2020]. Conveniently, SWOT will explicitly measure $S_h$ at unprecendeted spatial and temporal resolutions and will be spatially joined to hydrography that provides $S_0$. Thus, future researchers can use BIKER and SWOT in conjunction to directly explore this question.

@@

##6 Acknowledgements
C.B. Brinkerhoff was funded on xxxxxxxxxxxxx. BIKER is available at https://github.com/craigbrinkerhoff/BIKER. All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600. Data used in this study was made generously available by Frasson et al. (2019), Ulseth et al. (2019), and Beaulieu et al. (2012). This manuscript benefitted extensively from the decade plus body of work generated by the SWOT Discharge Algorithm Working Group.

>>>>

##Key Points (85 characters each)
- We introduce the BIKER algorithm to remotely sense riverine gas exchange using the SWOT satellite
- BIKER reproduces estimates of $CO_2$ fluxes when coupled with in situ gas concentration data
- Expected SWOT measurement errors exert a trivial influence on BIKER's estimates
- BIKER's accuracy is limited by process-level uncertainties and not by its algorithmic setup

@@

$$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})  \mathbf{(5)}$$

@@

##4 Discussion
###4.1 Towards remote sensing global spatiotemporal dynamics of $k_{600}$
To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g. @alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @schelkerCO2EvasionSteep2016a], 3) discharge [e.g. @robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998], or 4) oxygen diel curves (where gas exchange is still assumed constant in time- e.g. @applingOvercomingEquifinalityLeveraging2018; @demarsStreamMetabolismOpen2015). However, the spatiotemporal dynamics of gas exchange are still weakly constrained. A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012]. @wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating river specific (‘at-a-station’) temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed. This limited knowledge of large-scale $k_{600}$ spatiotemporal dynamics is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also a lack of measurements. Authors have argued that the key to explaining the large residual variation in upscaling models is to explore at-a-station temporal variability in $k_{600}$ [@hallGasExchangeStreams2020].

@@

###4.2 Estimating bulk carbon efflux using the SWOT satellite
Section 3.3 confirms that BIKER is successful at 1) reproducing $FCO_2$ without any in situ river information for $k_{600}$ (Figure 7) and 2) reproducing the bulk carbon efflux without any in situ information for $k_{600}$ (Figure 5.8). This encouraging result has two main implications for future work. First, section 2.4 confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers. This is particualry useful given recent advances in high temporal resoultion in situ gas concentration loggers **get citations from Kelly**. BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope following recent work doing the same using the [@hagemannBAMBayesianAMHGManning2017a] RSQ algorithm [@harlanDischargeEstimationDense].

Secondly, it is important to stress that unlike BIKER, the HG models in Figure 8 rely on an in situ streamgauge. This means that Figure 8 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease, though this decrease would not be too much due to the generally good performance of those data. Thus, these results suggest that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our bulk efflux understandings at gauged sites too. Future work should systematically quantify prediction error from coupling global-scope HG models with modeled discharge, as is the default workflow used in recent upscaling studies [e.g. @lauerwaldSpatialPatternsCO22015, @horgbyUnexpectedLargeEvasion2019]. **I assume Shaoda's paper will still be in review and can't cite here **

@@

###4.2 Sources of BIKER estimate uncertainit: process-level or remote sensing?
Throughout the BIKER validation, we have assumed no parameter uncertainity in the upscaling parameters $a_i$ and $b_i$. We have shown that BIKER estimate uncertainty is significantly less than $a_i$ and $b_i$ uncertainity (Figures 4 and 5). Therefore, BIKER provides similar estimates of $k_{600}$ to those produced by current upscaling models and most of the total uncertainity stems from the upscaling model itself (Figure 4). We argue that BIKER's estimate uncertainity is limited by current process-level knowledge of $k_{600}$ upscaling and not by BIKER or SWOT, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the physical processes governing gas exchange from SWOT-observable rivers. Developing and validating the BIKER algorithm has provided a unique lense in which we can probe current understandings of riverine gas exchange theory. In that context, we next explore some considerations for future work on $k_{600}$ upscaling in SWOT-observable rivers.

####4.2.1 Gas exchange under uniform and non-uniform flow conditions
Section 2.2 highlights that most upscaling studies to date have assumed uniform flow conditions (i.e. $S_h \neq S_0$) in order to train upscaling models using readily available slope data. However, the first-principles model previously used by [@ulsethDistinctAirWater2019a] and [@raymondScalingGasTransfer2012a] to define *eD* [@tsivoglouTracerMeasurementReaeration1976] does not make this simplifying assumption. Therefore, it is an open research question whether parameterizing $k_{600}$ upscaling models via $S_h$ can account for some of the unexplained residual variation in current upscaling models [@hallGasExchangeStreams2020]. Conveniently, SWOT will explicitly measure $S_h$ at unprecendeted spatial and temporal resolutions and will be spatially joined to hydrography that provides $S_0$. Thus, future researchers can use BIKER and SWOT in conjunction to directly explore this question.

@@

##6 Acknowledgements
C.B. Brinkerhoff was funded on xxxxxxxxxxxxx. BIKER is available at https://github.com/craigbrinkerhoff/BIKER. All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600. Data used in this study was made generously available by the authors of Frasson et al. (2019), Ulseth et al. (2019), and Beaulieu et al. (2012). This manuscript benefitted extensively from the decade plus body of work generated by the SWOT Discharge Algorithm Working Group.
